---
title: "Assignment 1 Task 1 Sierra Amphibians"
author: "Zoe Zhou"
format: 
  html:
    code-fold: true
    embed-resources: true
    toc: true
execute:
  warning: false
  message: false
theme: minty
---
## Introduction

This analysis examines population trends of the endangered mountain yellow-legged frog (*Rana muscosa*). Two visualizations were created: 

1. **Annual Trends by Life Stage**: A graph of yearly total counts of adults, subadults, and tadpoles across all water bodies (excluding egg masses). 

2. **Top 5 Lakes by Adult/Subadult Counts**: A bar chart showing total counts in the five lakes with the highest populations, excluding tadpoles and anonymizing lake IDs (e.g., “Lake 10025”). 

These graphs provide insights into population dynamics and key habitats for conservation.

## Dataset
**Amphibian Abundance Data**

Source: [The Sierra Lakes Inventory Project](https://portal.edirepository.org/nis/metadataviewer?packageid=edi.577.2)

The Sierra Lakes Inventory Project (SLIP) was a research endeavor that ran from 1995-2002 and has supported research and management of Sierra Nevada aquatic ecosystems and their terrestrial interfaces. This dataset described the physical characteristics of and surveyed aquatic communities for >8,000 lentic water bodies in the southern Sierra Nevada, including lakes, ponds, marshes, and meadows.


## Set-up
We will use the following libraries and set-up through this analysis

```{r}
# Import libraries
library(tidyverse)
library(tidymodels)
library(here)
library(cowplot)
library(readxl)
library(patchwork)
library(dplyr)
library(lubridate)
library(ggplot2)

```

## Load in Data
Use basic command to explore data
```{r}
#| fold: false
df <- read_excel('data/sierra_amphibians.xlsx')
#head(df)
summary(df)
```
## Part 1 
Provide a graph of total mountain yellow-legged frog (Rana muscosa) count each year across all water bodies, by life stage excluding the ‘EggMass’ level. 

Data Wranging Pseudocode 

- drop columns or select columns

- filter out EggMass in amphibian_life_stage

- select by species

- check for na values

- convert data to datetime

- group by year and life stage

- get counts function

- ungroup()

- plot by year on x, counts on y, color by life stages 

Preview cleaned dataframe
```{r}
#| fold: false
# Clean dataframe
df_ramu <- df %>% 
  select(amphibian_species, amphibian_life_stage, amphibian_number, survey_date, lake_id) %>% 
  drop_na() %>% 
  filter(amphibian_species == "RAMU", amphibian_life_stage != "EggMass") 

# Convert timeseries to  datetime variables
df_ramu$survey_date <- as.factor(lubridate::year(df_ramu$survey_date)) # factorize year column

# group by year to get count 
df_ramu_sum <- df_ramu %>% 
  group_by(survey_date, amphibian_life_stage) %>% 
  summarise(total_counts=sum(amphibian_number, na.rm= TRUE)) %>% 
  ungroup()

# Check data frame
knitr::kable(head(df_ramu_sum))
```
Plot histogram
```{r}
# Plot histograms
plot1 <- ggplot(data=df_ramu_sum, mapping = aes(x=survey_date, y=total_counts, fill=amphibian_life_stage)) +
  geom_col() + 
  labs(
    title = "Population of Rana muscosa by Year",
    x = "Survey Year",
    y = "Total Counts of Rana muscosa"
  )+
  theme_light()+
  theme(
    plot.title = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 11),
    panel.grid = element_blank(),
    legend.position = c(0.4, 0.8)
  )
print(plot1)
```
## Part 2 
Create a column graph containing total counts (over all years of the study) of combined adult and subadult endangered mountain yellow-legged frogs (Rana muscosa) observed in the 5 lakes with the greatest total observed counts. 

Data Wrangling
```{r}
#| fold: false
# Filter out tadpoles
df_ramu_adult <- filter(df_ramu, amphibian_life_stage != "Tadpole") 

# group by lake to get count 
ramu_lake <- df_ramu_adult %>% 
  group_by(lake_id) %>% 
  summarise(total_counts=sum(amphibian_number, na.rm= TRUE)) %>% 
  ungroup() %>% 
  arrange(desc(total_counts))

# Check top 5 lakes
knitr::kable(head(ramu_lake, 5))

```
Prepare for plotting
```{r}
#| fold: false
# Prepare for plotting
top_lakes <- ramu_lake %>%
  mutate(lake_label = paste("Lake", lake_id)) %>% 
  mutate(lake_label = fct_reorder(lake_label, total_counts, .desc = TRUE)) %>%  
  slice_head(n = 5)

# Sort by total_counts and factorize lake id
#ramu_lake$lake_id <- factor(ramu_lake$lake_id, #levels=ramu_lake$lake_id[order(-ramu_lake$total_counts)])
#top_lakes <- ramu_lake %>% 
#  slice_head(n=5)# %>% 
# print(top_lakes)
# Plot histograms
plot2 <- ggplot(data=top_lakes, mapping = aes(x=lake_label, y=total_counts)) +
  geom_col(fill = "steelblue") + 
  labs(
    title = "Rana muscosa Lake Population",
    x = "Lake ID",
    y = "Total Counts of RAMU"
  ) +
    geom_text(aes(label = total_counts), 
            vjust = -0.5)+
  lims(y = c(0, 2600))+
  theme_light()+
  theme(
    plot.title = element_text(face = "bold", size = 10),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10),
    axis.title = element_text(size = 11),
    panel.grid = element_blank()
  )
print(plot2)
```
Final Visualization
```{r}
#| fig-cap: Temporal and spatial distribution of Rana muscosa populations. Panel on the left shows the overall population trend across all surveyed lakes from 1995 to 2002. Panel on the right displays the five lakes with the highest recorded RAMU counts, highlighting significant spatial variation in population densities across different water bodies.

# Combine plots using patchwork
combined_plot <- plot1 + plot2 +
  plot_layout(widths = c(15, 15)) +
  plot_annotation(
    title = "Population Trends of Rana muscosa Across Sierra Lakes",
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )
  )

# Display combined plot
combined_plot

```

